---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Package - hclustHDCP

# Description 

This package deals with one of the classical problem in statistics, called change-point detection method. It attempts to detect abrupt significant changes in probability distributions for a sequence of observations arranged in chronological order. We implement change-point detection for high dimensional observations. This method detects change-points based on hierarchical clustering. The algorithm initially considers each observation as a cluster and at each subsequent step merges two closest consecutive clusters based on linkages (single, average or complete). Hence the number of clusters decreases by one in each step of the algorithm. In case the number of change-points to be detected is known, the algorithm stops when the desired number of clusters is reached. It returns the change-point locations where the cluster membership changes. However, if the number of change-points to detect is unknown, then we make use of some clustering evaluation algorithm like Penalized Dunn index to estimate optimal number of change-points. Here we also implement the algorithm using a different distance function, called the mean absolute deviation of distances (MADD distance) instead of usual Euclidean distance while calculating the linkage methods. This enables the method to perform well in low sample size situations as well for most of the occasions. Specifically when the changes in the observations are not in one dimensional marginal distributions and usual Euclidean distance fails to retain the neighbouhood structure of the observations.

# Installation Instruction
devtools::install_github("DawnTrisha/hclustHDCP")

# Use of the functions
As described earlier this package is intended to detect change-points specially in high dimensional low sample size regime when the changes are beyond mean change of the observations. There are four functions available.

## detect_single_cp : 
Detects single change-point in a data sequence arranged chronologically. Here are the points to consider while using the function

 - The user need to enter either the data matrix (X) or the distance matrix (D).

 - The function by default uses Euclidean distance to implement the algorithm.

 - If the one wants to use *genMADD distance* or *MADD distance*, use *distmat_HDLSS* function to first compute the distance matrix and then supply the matrix to the function *detect_single_cp* (See Example 2). 

```{r}
library(hclustHDCP)

##### Example 1
set.seed(1)
##### Generate data matrix
X1 = matrix(rnorm((15 * 50), mean = 0, sd = 1), nrow = 15, ncol = 50)
X2 = matrix(rnorm((15 * 50), mean = 4, sd = 1), nrow = 15, ncol = 50)
X = rbind(X1, X2)

detect_single_cp(X = X)  # Detect single change-point with default average linkage
detect_single_cp(X = X, dist.method = "single") # Detect single change-point with single linkage

##### Example 2
set.seed(1)
##### Generate data matrix
X1 = matrix(rnorm((15 * 50), mean = 0, sd = 1), nrow = 15, ncol = 50)
X2 = matrix(rnorm((15 * 50), mean = 4, sd = 1), nrow = 15, ncol = 50)
X = rbind(X1, X2)

##### Distance matrix (MADD distance)
D_mat = distmat_HDLSS(X = X, option = "MADD")

detect_single_cp(D = D_mat, dist.method = "complete")  # Detect single change-point with MADD distance matrix and complete linkage
```


## detect_multiple_cp : 
Detects multiple change-points in a data sequence arranged chronologically. Here also we need to consider the same three points that we used for detect_single_cp function. Additionally we also need to supply the number of change-points to detect (1 < numcp < n). Here is an illustration.

```{r}
##### Example 3
set.seed(1)
##### Generate data matrix
X1 = matrix(rnorm((15 * 50), mean = 0, sd = 1), nrow = 15, ncol = 50)
X2 = matrix(rnorm((15 * 50), mean = 1, sd = 1), nrow = 15, ncol = 50)
X3 = matrix(rnorm((15 * 50), mean = 2, sd = 1), nrow = 15, ncol = 50)
X = rbind(X1, X2, X3)

##### Detect two change-points with default average linkage
detect_multiple_cp(X = X, numcp = 2)

##### Distance matrix (genMADD distance)

D_mat = distmat_HDLSS(X = X, option = "MADD")

##### Detect two change-points with default average linkage
detect_multiple_cp(D = D_mat, numcp = 2)
```
