---
title: "Cluster Based High Dimensional Change-point Detection"
author: "Trisha Dawn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cluster Based High Dimensional Change-point Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hclustHDCP)
```
# Objective

In this vignette for the package **hclustHDCP** we will illustrate the importance of the method and the use of the package.

# Introduction

Change-point detection is a classical problem in statistics and machine learning which refers to detecting abrupt significant changes in the probability distribution of a sequence of observations arranged chronologically. There could be single change-point or multiple change-points in the sequence of observations where the observations may be univariate, multivariate, networks or even functional in nature. Clustering can be incorporated in an attempt to detect change-points because we can think of existence of clusters in the observation sequence based on the change-point locations. More precisely, if there is a change-point $1< \tau < n$ in an observed sequence of observations $\{x_1, x_2, \dots, x_n\}$, then there are two clusters, namely, $\{x_1, x_2, \dots, x_{\tau}\}$ and $\{x_{\tau + 1}, x_{\tau + 2}, \dots, x_n\}$. 

This package uses hierarchical clustering based on linkage methods to solve the problem of finding change-point locations. The following example illustrates this. This example deals with two change-points for two dimensional normal observations differing in their location vector before and after change-points.   

